Synthetic algebraic geometry and the phenomenon of nongeometric sequents

### 2018-04-05 slides-pssl103.tex
Slides for PSSL 103 (WIP)

>>>>

\documentclass[12pt,utf8,notheorems,compress,t]{beamer}
\usepackage{etex}

\usepackage[english]{babel}

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows,patterns}
\hypersetup{colorlinks=true}

\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

\graphicspath{{images/}}

\usepackage[protrusion=true,expansion=true]{microtype}

\setlength\parskip{\medskipamount}
\setlength\parindent{0pt}

\title{Synthetic algebraic geometry and the phenomenon of nongeometric sequents}
\author{Ingo Blechschmidt}
\date{April 7th, 2018}

\useinnertheme[shadow=true]{rounded}
\useoutertheme{split}
\usecolortheme{orchid}
\usecolortheme{whale}
\setbeamerfont{block title}{size={}}

\useinnertheme{rectangles}

\usecolortheme{seahorse}
\definecolor{mypurple}{RGB}{150,0,255}
\setbeamercolor{structure}{fg=mypurple}
\definecolor{myred}{RGB}{150,0,0}
\setbeamercolor*{title}{bg=myred,fg=white}
\setbeamercolor*{titlelike}{bg=myred,fg=white}

\usefonttheme{serif}
\usepackage[T1]{fontenc}
\usepackage{libertine}

\newcommand{\A}{\mathcal{A}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\DeclareMathOperator{\Spec}{Spec}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}

\setbeamertemplate{blocks}[rounded][shadow=false]

\newcommand{\fmini}[2]{\framebox{\begin{minipage}{#1}#2\end{minipage}}}
\makeatletter
\def\underunbrace#1{\mathop{\vtop{\m@th\ialign{##\crcr
      $\hfil\displaystyle{#1}\hfil$\crcr\noalign{\kern3\p@\nointerlineskip}
      \crcr\noalign{\kern3\p@}}}}\limits}
\def\overunbrace#1{\mathop{\vbox{\m@th\ialign{##\crcr\noalign{\kern3\p@}
      \crcr\noalign{\kern3\p@\nointerlineskip}
      $\hfil\displaystyle{#1}\hfil$\crcr}}}\limits}
\makeatother

\newenvironment{changemargin}[2]{%
  \begin{list}{}{%
    \setlength{\topsep}{0pt}%
    \setlength{\leftmargin}{#1}%
    \setlength{\rightmargin}{#2}%
    \setlength{\listparindent}{\parindent}%
    \setlength{\itemindent}{\parindent}%
    \setlength{\parsep}{\parskip}%
  }%
  \item[]}{\end{list}}

\newcommand{\pointthis}[2]{%
  \tikz[remember picture,baseline]{\node[anchor=base,inner sep=0,outer sep=0]%
    (#1) {#1};\node[overlay,rectangle callout,%
    callout relative pointer={(-0.5cm,1.0cm)},fill=blue!20] at ($(#1.north)+(0.9cm,-1.6cm)$) {#2};}%
}%

\newcommand{\hcancel}[5]{%
  \tikz[baseline=(tocancel.base)]{
    \node[inner sep=0pt,outer sep=0pt] (tocancel) {#1};
    \draw[red, line width=0.4mm] ($(tocancel.south west)+(#2,#3)$) -- ($(tocancel.north east)+(#4,#5)$);
  }%
}

\newcommand{\slogan}[1]{%
  \begin{center}%
    \setlength{\fboxrule}{2pt}%
    \setlength{\fboxsep}{8pt}%
    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal text}\parbox{0.93\textwidth}{#1}}}%
  \end{center}%
}

\newcommand{\sloganwithoutborder}[1]{%
  \begin{center}%
    \setlength{\fboxrule}{0pt}%
    \setlength{\fboxsep}{-14pt}%
    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal
    text}\parbox{0.9\textwidth}{\begin{center}#1\end{center}}}}%
  \end{center}%
}

% Adapted from https://latex.org/forum/viewtopic.php?t=2251 (Stefan Kottwitz)
\newenvironment<>{varblock}[2]{
  \begin{center}
    \begin{minipage}{#1}
      \setlength{\textwidth}{#1}
      \begin{actionenv}#3
	\def\insertblocktitle{\centering #2}
	\par
	\usebeamertemplate{block begin}}{
        \par
        \usebeamertemplate{block end}
      \end{actionenv}
    \end{minipage}
  \end{center}}

\setbeamertemplate{frametitle}{%
  \vskip0.7em%
  \leavevmode%
  \begin{beamercolorbox}[dp=1ex,center]{}%
      \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
  \end{beamercolorbox}%
}

\setbeamertemplate{navigation symbols}{}

\newcounter{framenumberpreappendix}
\newcommand{\backupstart}{
  \setcounter{framenumberpreappendix}{\value{framenumber}}
}
\newcommand{\backupend}{
  \addtocounter{framenumberpreappendix}{-\value{framenumber}}
  \addtocounter{framenumber}{\value{framenumberpreappendix}} 
}

\setbeamertemplate{footline}{%
  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex,dp=1ex,right,rightskip=1mm,leftskip=1mm]{}%
    \inserttitle \hfill
    \insertframenumber\,/\,\inserttotalframenumber
  \end{beamercolorbox}%
  \vskip0pt%
}

\newcommand{\hil}[1]{{\usebeamercolor[fg]{item}{\textbf{#1}}}}

\IfSubStr{\jobname}{\detokenize{nonotes}}{
  \setbeameroption{hide notes}
}{
  \setbeameroption{show notes}
}
\setbeamertemplate{note page}[plain]
\setbeameroption{hide notes}

\begin{document}

\addtocounter{framenumber}{-1}

\begin{frame}[c]
  \centering
  \includegraphics[scale=0.3]{images/external-internal-small}
  \medskip

  \hil{Synthetic algebraic geometry} \\
  \small\emph{a case study in applied topos theory}

  \large\&

  \normalsize

  \hil{the phenomenon of nongeometric sequents}

  \medskip

  \scriptsize
  Ingo Blechschmidt \\
  University of Augsburg
  \medskip

  103rd Peripatetic Seminar on Sheaves and Logic (Brno) \\
  April 7th, 2018
  \par
\end{frame}

\section{Preliminaries}

\subsection{Synthetic differential geometry}

\begin{frame}{Synthetic differential geometry}
  \vspace*{-1em}
  \begin{varblock}{9.5cm}{The axiom of microaffinity}
    \justifying
    Let~$\Delta = \{ \varepsilon \in \RR \,|\, \varepsilon^2 = 0 \}$.
    For any function $f : \Delta \to \RR$,
    there are unique numbers~$a, b \in \RR$ such that
    $f(\varepsilon) = a + b \varepsilon$
    for all~$\varepsilon \in \Delta$.
  \end{varblock}

  \begin{center}
    \includegraphics[width=0.5\textwidth]{microaffinity}
  \end{center}

  \vspace*{-1em}
  \begin{itemize}
    \item The \hil{derivative} of~$f$ as above at zero is~$b$.
    \item Manifolds are \hil{just sets}.
    \item A \hil{tangent vector} to~$M$ is a map~$\Delta \to M$.
  \end{itemize}

  \centering
  \hil{Toposes provide models for this theory.}
\end{frame}

\subsection{Toposes}

\begin{frame}{The internal universe of a topos}
  A \hil{topos} is a category which has finite limits,
  is cartesian closed and has a subobject classifier, for instance
  \begin{itemize}
    \item \hil{$\boldsymbol{\Set}$}, the category of sets;
    \item \hil{$\boldsymbol{\Sh(X)}$}, the category of set-valued sheaves over
    a space~$X$;
    \item \hil{$\boldsymbol{\Eff}$}, the ef{}fective topos (roughly: a
    category of data types).
  \end{itemize}

  \slogan{Any topos supports an \hil{internal language},
  which is sound with respect to
  \only<1>{\hil{intuitionistic reasoning}}%
  \only<2->{\hil{\pointthis{intuitionistic reasoning}{
    \textnormal{\textcolor{black}{
    no $\varphi \vee \neg\varphi$,\ \
    no $\neg\neg\varphi \Rightarrow \varphi$,\ \
    no axiom of choice}}}}}.}
\end{frame}

\newcommand{\intex}[3]{#1 \qquad #2\par #3\medskip}

\begin{frame}{Curious universes}
  \small
  \intex{
    $\Eff \models \text{``There are infinitely many prime numbers.''}$
  }{\textcolor{green!90}{\cmark}}{
    External meaning: There is a (Turing) \hil{machine} producing arbitrarily many prime
    numbers.
  }

  \intex{
    $\Eff \not\models \text{``Any Turing machine halts or doesn't halt.''}$
  }{\textcolor{red!80}{\xmark}}{
    External meaning: There is a \hil{halting oracle} which determines whether
    any given machine halts or doesn't halt.
  }

  \intex{
    $\Sh(X) \not\models \text{``Any cont. function with opposite signs has
    a zero.''}$
  }{\textcolor{red!80}{\xmark}}{
    External meaning: Zeros can be picked \hil{locally continuously} in
    continuous families of continuous functions.
  }
\end{frame}

\end{document}
